---
title: "July Temperature Trend — Willamette River"
author: "Katerina Bischel"
date: today
format: html
---

## Overview

This document analyzes whether July water temperatures at three Willamette River
USGS sites show a significant upward or downward trend from 2010–2023, and whether
that trend differs across sites. A GLM with a year × site interaction is used to
estimate the slope (°C per year) at each site and test whether it is statistically
significant.

```{r}
#| message: false
#| warning: false

# tidyverse provides dplyr for filtering/summarising and ggplot2 for visualization
library(tidyverse)
```

## Load Data

```{r}
# load the locally saved CSV produced by willamette_temperature_data.qmd
# this avoids re-downloading from the NWIS server
temp_data <- read_csv("data/willamette_temp.csv")
```

## GLM Trend Analysis

```{r}
# filter the full dataset down to July observations only
july_data <- subset(temp_data, month == "Jul")

# fit a GLM to test for a linear trend in July temperatures over time, separated by site
# year*site_no includes: year, site_no, and the year:site_no interaction
# the interaction allows each site to have a different slope (warming rate)
july_model <- glm(temp_C ~ year*site_no, data = july_data)

# print the full model summary including coefficients, standard errors, and p-values
summary(july_model)
```

## Interpretation

The GLM output tells us:

- **`year` (Estimate)**: the slope (°C/year) for the reference site (14166000 — first alphabetically). A positive value means temperatures are increasing over time at that site.
- **`year:site_no14211010` and `year:site_no14211720`**: how much steeper (if positive) or shallower (if negative) the warming trend is at those sites *relative* to the reference site. Add these to the `year` estimate to get each site's actual slope.
- **`(Intercept)`, `site_no14211010`, `site_no14211720`**: set the baseline temperature to make the regression work — not used for trend interpretation. Mean temperatures per site are better compared using simple means.

Based on the model output, all three sites show a **positive (warming) trend** in July temperatures. The `year` coefficient is statistically significant (p < 0.001), indicating a meaningful increase in temperature over the study period.

## Trend Visualization

```{r}
# calculate the mean July temperature per year and site for plotting
# this smooths out day-to-day variation within the month
july_annual <- july_data |>
  group_by(year, site_no) |>
  summarise(mean_temp = mean(temp_C, na.rm = TRUE), .groups = "drop")

# plot annual mean July temperatures with a linear trend line per site
# shaded band shows the 95% confidence interval around each trend line
ggplot(july_annual, aes(x = year, y = mean_temp, color = site_no)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "July Mean Stream Temperature Trend (2010–2023)",
    x = "Year",
    y = "Mean Temperature (°C)",
    color = "Site Number"
  ) +
  theme_minimal()
```
