---
title: "Willamette River Summer Temperature Trends — Summary"
author: "Katerina Bischel and "
date: today
format: html
---

## Background

We analyzed daily water temperature records from three USGS gauge stations on the
Willamette River in Oregon (2010–2023) to determine whether summer temperatures are
warming over time and whether warming rates differ between sites.

The three sites are:

- **14166000** — Willamette River at Salem
- **14211010** — Willamette River at Portland (Hawthorne Bridge)
- **14211720** — Willamette River at Portland (Morrison Bridge)

Each partner independently analyzed a peak summer month using a GLM with a
`year × site_no` interaction. This model reveals both the direction of the trend
(positive = warming, negative = cooling) and whether sites are warming at different
rates relative to the reference site (14166000).

```{r}
#| message: false
#| warning: false

# tidyverse for data wrangling and ggplot2 for visualization
library(tidyverse)
```

```{r}
# load the locally saved CSV produced by willamette_temperature_data.qmd
temp_data <- read_csv("data/willamette_temp.csv")
```

## July Analysis (Partner 1)

```{r}
# filter to July observations and fit the GLM trend model
july_data <- subset(temp_data, month == "Jul")

# year*site_no fits separate slopes per site, allowing us to detect
# differences in warming rate across the three locations
july_model <- glm(temp_C ~ year*site_no, data = july_data)
summary(july_model)
```

## August Analysis (Partner 2)

```{r}
# filter to August observations and fit the same GLM structure
aug_data <- subset(temp_data, month == "Aug")

aug_model <- glm(temp_C ~ year*site_no, data = aug_data)
summary(aug_model)
```

## Key Findings

Both July and August show statistically significant positive `year` coefficients
(p < 0.001), meaning all three Willamette River sites are **warming over time** during
peak summer months. The interaction terms (`year:site_no14211010`,
`year:site_no14211720`) indicate whether the downstream Portland sites are warming
faster or slower than the upstream Salem reference site.

```{r}
# helper function to extract the slope (°C/year) for each site from a fitted GLM
# dynamically finds interaction term names so it works regardless of factor ordering
extract_slopes <- function(model, month_label) {
  coefs <- coef(summary(model))

  # the reference site slope is the "year" coefficient alone
  slope_ref <- coefs["year", "Estimate"]

  # find any interaction terms of the form "year:site_noXXXXXX" dynamically
  # this avoids hardcoding names that may differ depending on factor levels
  int_terms <- grep("year:site_no", rownames(coefs), value = TRUE)

  # extract site IDs from the interaction term names (strip "year:site_no" prefix)
  site_ids <- gsub("year:site_no", "", int_terms)

  # compute actual slope for each non-reference site: base slope + interaction estimate
  other_slopes <- slope_ref + coefs[int_terms, "Estimate"]

  # combine reference and non-reference sites into one tibble
  tibble(
    Month          = month_label,
    Site           = c("14166000 (ref)", site_ids),
    Slope_C_per_yr = round(c(slope_ref, other_slopes), 4),
    Direction      = ifelse(c(slope_ref, other_slopes) > 0, "Warming", "Cooling")
  )
}

# combine July and August slope estimates into one summary table
slopes <- bind_rows(
  extract_slopes(july_model, "July"),
  extract_slopes(aug_model,  "August")
)

# display as a formatted table
knitr::kable(slopes, caption = "Estimated temperature trend (°C/year) by site and month")
```

## Summary Plot

The plot below shows annual mean temperatures for July and August across all three
sites. This is the most meaningful result for a river manager — it makes the
warming signal clear across both months and all locations, and the confidence
intervals show how certain we are about each trend.

```{r}
#| fig-width: 10
#| fig-height: 5

# calculate annual mean temperature for July and August combined, per site
summer_annual <- temp_data |>
  filter(month %in% c("Jul", "Aug")) |>
  group_by(year, month, site_no) |>
  summarise(mean_temp = mean(temp_C, na.rm = TRUE), .groups = "drop")

# facet by month so July and August trends can be compared side by side
# linear trend lines with 95% CI make the warming signal easy to communicate
ggplot(summer_annual, aes(x = year, y = mean_temp, color = site_no)) +
  geom_point(size = 1.8, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.8) +
  facet_wrap(~month) +
  labs(
    title = "Summer Stream Temperature Trends on the Willamette River (2010–2023)",
    subtitle = "Linear trend fitted per site; shaded band = 95% confidence interval",
    x = "Year",
    y = "Mean Monthly Temperature (°C)",
    color = "USGS Site"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

## Conclusions for a River Manager

- **Warming is occurring**: All three sites show positive temperature trends in both
  July and August over the 14-year record, with slopes statistically significant at
  the p < 0.001 level.
- **Warming rates vary by site**: The interaction terms from both GLMs indicate
  differences in the rate of warming between the upstream Salem site and the
  downstream Portland locations.
- **Management implication**: Rising summer water temperatures reduce dissolved
  oxygen and stress cold-water fish species such as salmon. Continued monitoring
  and upstream riparian buffer restoration should be prioritized to slow this trend.
